---
title: "Bibentry for FAIR datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bibentry for FAIR datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dataset)
iris_dataset_2 <- iris_dataset
```

An important aim of the **dataset** package is to create native R objects where bibliographic metadata cannot be detached, thus ensuring *Findability*, *Accessibility*, *Interoperability* and *Reusability* in the long run. We provide an interface and methods to add metadata required by open data repositories according to the more general [DCTERMS](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/) (Dublin Core) library metadata standard, or the more specific [DataCite](https://datacite-metadata-schema.readthedocs.io/en/4.6/) metadata standard.

The R statistical environment and language comes with several built-in datasets that were used by generations of users to learn the language or statistics itself. For the purposes of demonstrating the increased interoperability of our `dataset_df()` objects, the `iris` dataset is the most suitable from R's built in datasets.

```{r bibentryprint}
print(get_bibentry(iris_dataset_2), "Bibtex")
```
In 1935, American botanist Edgar Anderson measured petal and sepal structural dimensions (length and width) for 50 flowers from three Iris species: *Iris setosa*, *Iris versicolor*, and *Iris virginica*. 

The manageable but non-trivial size (5 variables and 150 total observations) and characteristics of Anderson’s Iris dataset, including linear relationships and multivariate normality, have made it amenable for introducing a wide range of statistical methods including data wrangling, visualization, linear modeling, multivariate analyses, and machine learning. 

The Iris dataset is built into a number of software packages including the auto-installed datasets package in R, Python’s scikit-learn machine learning library, and the SAS Sashelp library. As a result, eighty-six years after the data were initially published, the Iris dataset remains ubiquitous in statistics, computational methods, software documentation, and data science courses and materials, even though it is criticised for poor metadata, and because a R. A. Fisher referred to it fist in statistics. (You can read more about this controversy at the end of this vignette article.)

## Titles

To get the title of a dataset: 

```{r, get-title}
dataset_title(iris_dataset)
```
To change the title: 

```{r, change-title}
dataset_title(iris_dataset_2, overwrite = TRUE) <- "The Famous Iris Dataset"
get_bibentry(iris_dataset_2)
```
## Creators

The \code{Creator} corresponds to [dct:creator](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/elements11/creator/)
in Dublin Core and Creator in DataCite, the two most important metadata definitions for publishing datasets in repositories. They refer to the The name of the entity that holds, archives, publishes prints, distributes, releases, issues, or produces the dataset. This property will be used to
formulate the citation.

```{r creator}
creator(iris_dataset)
```
```{r creator-modify}
iris_dataset_2 <- iris_dataset
# Add a new creator, with overwriting existing authorship information:
creator(iris_dataset_2, overwrite = TRUE) <- person("Jane", "Doe", role = "aut")

# Add a new creator, without overwriting existing authorship information:
creator(iris_dataset_2, overwrite = FALSE) <- person("John", "Doe", role = "ctb")

# The two new creation contributors:
creator(iris_dataset_2)
```


## Further descriptive metadata about the whole dataset

### Publication year

The publication year is usually one of the most important descriptive metadata in repositories and libraries:

```{r irispublicationyear}
publication_year(iris_dataset_2)
```

### Language

```{r, language}
# Get the language:
language(iris_dataset)

# Reset the language:
language(iris_dataset_2) <- "French"
language(iris_dataset_2)
```
### Rights statement


```{r rights}
# Add rights statement to the dataset
rights(iris_dataset_2, overwrite = TRUE) <- "GNU-2"
```

Some metadata functions prevent accidental overwriting, except for the default `:unas` unassigned and `:tba` to-be-announced values.

```{r prevent-overwrite}
rights(iris_dataset_2) <- "CC0"
rights(iris_dataset_2)
```

Overwriting the rights statement needs an explicit approval:

```{r overwrite}
rights(iris_dataset_2, overwrite = TRUE) <- "GNU-2"
```

DataCite currently allows the use of subproperties. For example, the _Creative Commons Attribution 4.0 International_ would be described as: 

```{r rightsubproperties}
list(
  schemeURI = "https://spdx.org/licenses/",
  rightsIdentifierScheme = "SPDX",
  rightsIdentifier = "CC-BY-4.0",
  rightsURI = "https://creativecommons.org/licenses/by/4.0/"
)
```

The use of subproperties will be later implemented.

### Description

The description is currently implemented as a character string.  However, 
DataCite 4.6 states that if Description is used, `descriptionType` is mandatory.
This will be implemented later.

```
<descriptions>
    <description xml:lang="en" descriptionType="Abstract">Example abstract</description>
</descriptions>
```
```{r printdescription}
description(iris_dataset)
```


### Subject

```{r subject}
subject(iris_dataset)
```


```
<subjects>
  <subject xml:lang="en" subjectScheme="Library of Congress Subject Headings (LCSH)" schemeURI="https://id.loc.gov/authorities/subjects.html" valueURI="https://id.loc.gov/authorities/subjects/sh2009009655.html">Climate change mitigation</subject>
  <subject xml:lang="en" subjectScheme="ANZSRC Fields of Research" schemeURI="https://www.abs.gov.au/statistics/classifications/australian-and-new-zealand-standard-research-classification-anzsrc" classificationCode="370201">Climate change processes</subject>
</subject>
```

```{r subjectcreate}
subject_create(
  term = "data sets",
  subjectScheme = "Library of Congress Subject Headings (LCSH)",
  schemeURI = "https://id.loc.gov/authorities/subjects.html",
  valueURI = "http://id.loc.gov/authorities/subjects/sh2018002256"
)
```


### Identifiers

```{r identifier}
# Add rights statement to the dataset
identifier(iris_dataset_2)
```
## All bibliographic information

Get the metadata according to the DataCite definition:

```{r datacite}
print(as_datacite(iris_dataset), "Bibtex")
```

And according to DCTERMS (Dublin Core):


```{r dc}
library(RefManageR)
print(as_dublincore(iris_dataset), "Bibtex")
RefManageR::toBiblatex(as_dublincore(iris_dataset))
```

## The iris dataset controversy

Millions of R users, and countless students of statistics who do not use R are familiar with the iris dataset, which is about flowers; however, it is entirely appropriate to criticise the use of this dataset due to the fact that for many years Ronald Fisher, who introduced `iris` into statistical education, promoted his eugenic ideas through his activity as a member of the Eugenics Society and its council.

The data set consists of 50 samples from each of three species of Iris (Iris setosa, Iris virginica and Iris versicolor). Four features were measured from each sample: the length and the width of the sepals and petals, in centimeters. Based on the combination of these four features, Fisher developed a linear discriminant model to distinguish each species. Fisher's paper was published in the _Annals of Eugenics_ (today the _Annals of Human Genetics_). The journal obtained its current name in 1954 to reflect changing perceptions on eugenics. 

This study cemented the use of this dataset in the education of statistics; in the end of this vignette, you can read more about a careful analysis of the somewhat troubled intellectual heritage of statistician who introduced the dataset into the education of statistics.

In their R Journal article, Allison M. Horst, Alison Presmanes Hill, and Kristen B. Gorman: [Palmer Archipelago Penguins Data in the palmerpenguins R Package - An Alternative to Anderson’s Irises](https://journal.r-project.org/articles/RJ-2022-020/) criticise the dataset for lacking data documentation, metadata, and various other aspects, including, in their view that the dataset "is burdened by a history in eugenics research."

Walter Bodmer, R. A. Bailey, Brian Charlesworth, Adam Eyre-Walker, Vernon Farewell, Andrew Mead & Stephen Senn: [The outstanding scientist, R.A. Fisher: his views on eugenics and race](https://www.nature.com/articles/s41437-020-00394-6). The purpose of this article "is neither to defend nor attack Fisher’s work in eugenics and views on race, but to present a careful account of their substance and nature." The authors conclude that recent criticism of R. A. Fisher concentrates "on very limited aspects of his work and focusses attention on some of his views, both in terms of science and advocacy. This is entirely appropriate, but in re-assessing his many contributions to society, it is important to consider all aspects, and to respond in a responsible way—we should not forget any negative aspects, but equally not allow the negatives to completely overshadow the substantial benefits to modern scientific research."

The author of this package does not endorse eugenics in any form, but also does not share the view that the `iris` dataset is burdened by a history of eugenics, because it was created by the botanist Edgar Anderson and it did not support Fisher's view on eugenics. The `iris` dataset for most students of statistics represents Fisher's undeniable contribution to the science of statistics and its education. On a more substantial note, the dataset offers more  semantics than the other built-in R datasets, is indeed poor in metadata to demonstrate all the features of the `dataset_df()` class, and this vignette may be rewritten in the future with a more suitable dataset.
