---
title: "Modernising Citation Metadata in R: Introducing `bibrecord`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bibrecord}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupvignette, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dataset)
```

## Why We Needed `bibrecord`

The base R function `utils::bibentry()` provides a way to structure citation metadata. While effective for simple references, it falls short for modern metadata standards like:

-   **Dublin Core Terms (DCTERMS)**

-   **DataCite**

Both require clear distinction between creators, contributors, and richly typed relationships — things `bibentry()` doesn't natively support.

To bridge this gap while preserving compatibility with base R, the `bibrecord` class builds on top of `bibentry`, adding support for:

-   Multiple `person()` entries for contributors

-   Metadata fields aligned with DCTERMS and DataCite

-   Safe serialization and printing

## What is `bibrecord`

A `bibrecord` is a standard `bibentry` object with additional fields stored in attributes. This means:

-   You can use it with any code expecting a `bibentry`

-   You gain structured metadata fields like `contributor`, `subject`, `identifier`

-   Methods like `print()` are extended to reflect these enrichments

## Creating a `bibrecord`

```{r bibrecord}
person_jane <- person("Jane", "Doe", role = "cre")
person_alice <- person("Alice", "Smith", role = "dtm")

rec <- bibrecord(
  title = "GDP of Small States",
  author = list(person_jane),
  contributor = list(person_alice),
  publisher = "Tinystat",
  identifier = "doi:10.1234/example",
  date = "2023-05-01",
  subject = "Economic indicators"
)
```


## Printing a `bibrecord`

```{r print}
print(rec)
```

This prints both the base citation and the contributors, clearly labeled.

## Compatibility with Existing Infrastructure

Because `bibrecord` inherits from `bibentry`:

-   It works with `citation()` and other base R citation tools

-   It integrates with existing bibliographic pipelines

-   You can convert it to `as_dublincore()` or `as_datacite()` without loss of structure

## Future Extensions

The `bibrecord` class can be further enriched by:

-   Adding support for `funder`, `geolocation`, `relatedIdentifier`, etc.

-   Exporting to JSON-LD or RDF formats

-   Integrating with Zenodo, Crossref, or Wikidata APIs

## Summary

| Feature        | `bibentry` | `bibrecord` |
|----------------|------------|-------------|
| Base R support | ✅         | ✅          |
| Contributors   | ❌         | ✅          |
| Semantic roles | ❌         | ✅          |
| DCTERMS export | ❌         | ✅          |
| DataCite-ready | ❌         | ✅          |

`bibrecord` is your drop-in upgrade for structured metadata in R.

Start simple. Stay compatible. Go semantic.
