---
title: "dataset_df: Create Datasets that are Easy to Share Exchange and Extend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dataset_df: Create Datasets that are Easy to Share Exchange and Extend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup}
library(dataset)
data("gdp")
```

The `dataset_df()` function helps create semantically rich datasets that are easier to:
- understand by humans,
- validate and process by machines,
- and share across tools, teams, and domains.

This vignette demonstrates how to construct a semantically enriched dataset using a subset of the  *GDP and main aggregates - international data cooperation annual data* dataset created by Eurostat (DOI: <https://doi.org/10.2908/NAIDA_10_GDP>).


```{r printgdp}
print(gdp)
```

While the raw dataset is valid, it can be hard to interpret without external documentation. For example:
- Countries are encoded in the `geo` variable.
- Reporting frequency (e.g., `A` for annual) is stored in `freq`

## Getting Started
The `dataset_df()` constructor enables two levels of semantic annotation for a tibble `tbl_df` object (created from a data.frame or tibble):
- Per-variable metadata: label, unit, definition, namespace
- Whole-dataset metadata: title, author, license, etc.

Let’s create a semantically enriched subset:

```{r createdataasetdf}
small_country_dataset <- dataset_df(
  geo = defined(
    gdp$geo,
    label = "Country name",
    concept = "http://purl.org/linked-data/sdmx/2009/dimension#refArea", 
    namespace = "https://dd.eionet.europa.eu/vocabulary/eurostat/geo/$1"), 
  year = defined(
    gdp$year, 
    label = "Reference Period (Year)", 
    concept = "http://purl.org/linked-data/sdmx/2009/dimension#refPeriod"),
  gdp = defined(
    gdp$gdp, 
    label = "Gross Domestic Product", 
    unit = "CP_MEUR", 
    concept = "http://data.europa.eu/83i/aa/GDP"),
  unit = defined(
    gdp$unit,
    label = "Unit of Measure", 
    concept  = "http://purl.org/linked-data/sdmx/2009/attribute#unitMeasure", 
    namespace = "https://dd.eionet.europa.eu/vocabulary/eurostat/unit/$1"
  ),
  freq = defined(
    gdp$freq, 
    label = "Frequency", 
    concept = "http://purl.org/linked-data/sdmx/2009/code"
  ),
  dataset_bibentry = dublincore(
    title = "Small Country Dataset", 
    creator = person("Jane", "Doe"), 
    publisher = "Example Inc.", 
    datasource = "https://doi.org/10.2908/NAIDA_10_GDP", 
    rights = "CC-BY", 
    coverage = "Andorra, Lichtenstein and San Marino")
  )
```

## Inspecting Variable-Level Metadata

The `defined` class columns retain the important semantic information about the label, exact concept definition link, and the unit of measure:

```{r varlabel}
var_label(small_country_dataset$gdp)
```
```{r varunit}
var_unit(small_country_dataset$gdp)
```
## Adding Dataset-Level Metadata

The `dataset_df()` object can also store metadata about the dataset as a whole. It follows widely adopted metadata standards:

- Dublin Core Terms (`dublincore()`), used in libraries and many data repositories
- DataCite (`datacite()`), a metadata definition mainly used in data repositories. 

Each metadata field can be accessed and modified using simple assignment functions.

For example, to specify the dataset language:

```{r language}
language(small_country_dataset) <- "en"
```

## Reviewing Dataset-Level Metadata
To inspect the full dataset description, use:

```{r bibentry}
print(get_bibentry(small_country_dataset), "bibtex" )
```
This prints a complete BibTeX-style entry, suitable for citation or export.

## Conclusion
With `dataset_df()`, your datasets become:

- Self-descriptive: variables carry labels, units, and definitions.
- Machine-readable: linked vocabularies and standard identifiers.
- Ready to publish and share: through metadata standards like Dublin Core and DataCite.

This approach supports FAIR data principles (Findable, Accessible, Interoperable, Reusable), and simplifies data reuse, interpretation, and validation.

Our package is aiming to work seamlessly with the rOpenSci [rdflib](https://github.com/ropensci/rdflib) package.  

