---
title: "defined"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{defined}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dataset)
```

`defined()` is a vector subclass of labelled. Labelled improves the semantic capacity of a base R factor with improved value levels and labels by adding a long-form, human-readable label to the variable itself.

```{r definedvector}
gdp_1 <- defined(
    c(3897, 7365), 
    label = "Gross Domestic Product", 
    unit = "million dollars", 
    definition = "http://data.europa.eu/83i/aa/GDP")
```

The `defined()` class extends the attributes of a labelled vector with a unit (of measure), a definition and a namespace. 

```{r definedattributes}
attributes(gdp_1)
cat("Get the label only: ")
var_label(gdp_1)
cat("Get the unit only: ")
var_unit(gdp_1)
cat("Get the definition only: ")
var_definition(gdp_1)
```
What happens if we try to concatenate a semantically under-specified new vector to the GDP vector?

```{r}
gdp_2 <- defined(2034, label = "Gross Domestic Product")
```

You will get an intended error message that some attributes are not compatible. You certainly want to avoid that you are concatenating figures in euros and dollars, for example.

```{r, eval=FALSE}
c(gdp_1, gdp_2)
Error in `vec_c()`:
! Can't combine `..1` <haven_labelled_defined> and `..2` <haven_labelled_defined>.
✖ Some attributes are incompatible.
```

Let's define better the GDP of San Marino:

```{r gpd2}
var_unit(gdp_2) <- "million dollars"
```


```{r vardef2}
var_definition(gdp_2) <- "http://data.europa.eu/83i/aa/GDP"
```

```{r c}
summary(c(gdp_1, gdp_2))
```
```{r country}
country = defined(c("AD", "LI", "SM"), 
                  label = "Country name", 
                  definition = "http://data.europa.eu/bna/c_6c2bb82d", 
                  namespace = "https://www.geonames.org/countries/$1/")
```

The point of using a namespace is that it can point to a both human- and machine readable definition of the ID column, or any attribute column in the datasets. (Attributes in a statistical datasets are characteristics of the observations or the measured variables.)

For example, the namespace definition above points to <https://www.geonames.org/countries/AD/> in the case of Andorra, <https://www.geonames.org/countries/LI/> for Lichtenstein, and <https://www.geonames.org/countries/SM/> for San Marino. And <http://publications.europa.eu/resource/authority/bna/c_6c2bb82d> resolves to a machine-readable definition of geographical names.

## Coerce to base R types

Coerce back the labelled country vector to a character vector:

```{r coerce-char}
as_character(country)
as_character(c(gdp_1, gdp_2))
```
And to numeric:

```{r coerce-num}
as_numeric(c(gdp_1, gdp_2))
```
