---
title: "Semantically Enriched Vectors with `defined()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Semantically Enriched Vectors with `defined()`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setupdefinedvignette, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `defined()` function creates semantically rich labelled vectors. These are a inherited from haven::labelled and extend the concept by adding:

- A human-readable variable label (available in haven::labelled)
- A unit of measurement
- A definition URI linking to external ontologies or concepts
- (Optionally) a namespace for controlled vocabularies

This structure enhances clarity and semantic precision, particularly useful for statistical metadata, time series, and linked open data.

## Getting Started

```{r setup}
library(dataset)
data("gdp")
```

We’ll start by wrapping a numeric GDP vector using `defined()`:

```{r gdp1}
gdp_1 = defined(
    gdp$gdp, 
    label = "Gross Domestic Product", 
    unit = "CP_MEUR", 
    concept = "http://data.europa.eu/83i/aa/GDP")
```

The `defined()` class builds on labelled vectors by adding rich metadata:

- label: a description of what the variable represents.
- unit: a short code for the measurement unit (e.g., CP_MEUR).
- definition: a URI linking to a concept or standard.
- namespace: (optional) for classifying coded values (e.g., SDMX codes).

This is particularly useful for reproducible research, standard-compliant data, or long-term interoperability. The class is defined with the attributes() function of R, and this guarantees a very wide compatibility. A defined vector can be used even in base R. 

```{r seeattributes}
attributes(gdp_1)
```
Use the `var_label()`, `var_unit()` and `var_defintion()` helper functions to set or retrieve metadata individually:

```{r convenience}
cat("Get the label only: ", var_label(gdp_1), "\n")
cat("Get the unit only: ", var_unit(gdp_1), "\n")
cat("Get the concept definition only: ", var_concept(gdp_1), "\n")
cat("All attributes:\n")
```
## Printing and Summary

```{r printdefined}
print(gdp_1)
```
```{r summarydefined}
summary(gdp_1)
```
## Handling ambiguity

What happens if we try to concatenate a semantically under-specified new vector to the GDP vector?

```{r ambiguous}
gdp_2 <- defined(
  c(2523.6, 2725.8, 3013.2), 
  label = "Gross Domestic Product")
```

You will get an intended error message that some attributes are not compatible. You certainly want to avoid that you are concatenating figures in euros and dollars, for example.

```{r, eval=FALSE}
c(gdp_1, gdp_2)
Error in `vec_c()`:
! Can't combine `..1` <haven_labelled_defined> and `..2` <haven_labelled_defined>.
✖ Some attributes are incompatible.
```

Let's define better the GDP of the Faroe Islands:

```{r gpd2}
var_unit(gdp_2) <- "CP_MEUR"
```


```{r vardef2}
var_concept(gdp_2) <- "http://data.europa.eu/83i/aa/GDP"
```

```{r c}
new_gdp <- c(gdp_1, gdp_2)
summary(new_gdp)
```
```{r country}
country = defined(
  c("AD", "LI", "SM"), 
  label = "Country name", 
  concept = "http://dd.eionet.europa.eu/vocabulary/eurostat/geo/", 
  namespace = "https://www.geonames.org/countries/$1/")
```

The point of using a namespace is that it can point to a both human- and machine readable definition of the ID column, or any attribute column in the datasets. (Attributes in a statistical datasets are characteristics of the observations or the measured variables.)

For example, the namespace definition above points to <https://www.geonames.org/countries/AD/> in the case of Andorra, <https://www.geonames.org/countries/LI/> for Lichtenstein, and <https://www.geonames.org/countries/SM/> for San Marino. And <http://publications.europa.eu/resource/authority/bna/c_6c2bb82d> resolves to a machine-readable definition of geographical names.

The use of several defined vectors in a dataset_df class explained in a separate vignette.

## Basic Usage

Working with character vectors:

```{r characters}
countries <- defined(
  c("AD", "LI"),
  label = "Country code",
  namespace = "https://www.geonames.org/countries/$1/"
)

countries
as_character(countries)
```

### Subsetting and coercion

```{r basicmethods}
gdp_1[1:2]
gdp_1 > 5000
as.vector(gdp_1)
as.list(gdp_1)
```
## Coerce to base R types

Coerce back the labelled country vector to a character vector:

```{r coerce-char}
as_character(country)
as_character(c(gdp_1, gdp_2))
```

And to numeric:

```{r coerce-num}
as_numeric(c(gdp_1, gdp_2))
```
