---
title: "Datasets With FAIR Metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Datasets With FAIR Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupknitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dataset)
```

## FAIR metadata

Our `dataset` R package aims to help the Findability, Accessibility, Interoperability, and Reuse of digital assets, particularly datacubes and datasets used in statistics and data analysis. The [FAIR](https://www.go-fair.org/fair-principles/) "…emphasise machine-actionability (i.e., the capacity of computational systems to find, access, interoperate, and reuse data with none or minimal human intervention) because humans increasingly rely on computational support to deal with data as a result of the increase in volume, complexity, and creation speed of data."

The `dataset` package adds metadata to R objects for full compatibility with the [Dublin Core Metadata Elements 1.1](https://www.dublincore.org/specifications/dublin-core/dces/), the [DataCite Mandatory Properties 4.4](https://support.datacite.org/docs/datacite-metadata-schema-44) and the [DataCite Recommended and Optional Properties 4.3](https://support.datacite.org/docs/schema-optional-properties-v43).

The `Dublin Core` is an ISO Standard 15836:2009 since February 2009 [ISO15836]. Currently the 15 "core" elements are part of a bigger [DCMI Metadata Terms](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/).



```{r createdataset}
iris_ds <- datacite_add(iris,
                        Title = "Iris Dataset",
                        Creator = person(family ="Anderson", given ="Edgar", role = "aut"),
                        Identifier = "https://doi.org/10.1111/j.1469-1809.1936.tb02137.x",
                        Publisher= "American Iris Society",
                        PublicationYear = 1935,
                        Geolocation = "US",
                        Language = "en", 
                        Version = "1.0")

datacite(iris_ds)
```

Currently, the DataCite properties (all mandatory, and what was filled up from the optional ones) can be seen along with standard R metadata. As a property of the `dataset` class, which follows W3C and SDMX standards, and `survey`, which will follow DDI standards, the history of the `dataset` from creation (import) will be recorded as attribute metadata.

```{r showattributes}
dct_iris <- dublincore_add(
  x = iris,
  Title = "Iris Dataset",
  Creator = person("Edgar", "Anderson", role = "aut"),
  Publisher = "American Iris Society",
  Identifier = "https://doi.org/10.1111/j.1469-1809.1936.tb02137.x",
  Date = 1935,
  Language = "en"
)

dublincore(dct_iris)
```
## FAir: Findable & Accessible Datasets

If you work with R, you are almost certainly familiar with the iris dataset.  The *?iris* will provide you with some information about this often used dataset in tutorials. But how you make sure that you do not forget its important properties?

The function datacite [DataCite](https://datacite.org/) add at least the mandatory properties of the [DataCite Metadata Schema 4.3](https://schema.datacite.org/), a list of core metadata properties chosen for an accurate and consistent identification of a resource for citation and retrieval purposes. DataCite is largely interoperable to the other similar international standard, the [Dublin Core](https://www.dublincore.org/).  We will later add similar `dublincore` function, however, the practical differences are so small that adjustments, if needed, can be easily made by hand.

```{r datacite}
library(dataset)
iris_dataset <- datacite_add(
  x = iris,
  Title = "Iris Dataset",
  Creator = person("Anderson", "Edgar", role = "aut"),
  Publisher= "American Iris Society",
  Identifier = "https://doi.org/10.1111/j.1469-1809.1936.tb02137.x",
  PublicationYear = 1935,
  Description = "This famous (Fisher's or Anderson's) iris data set gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica.",
  Language = "en")
```
```{r}
datacite(iris_dataset)
```
The `x` parameter can be any well-structured R object that meets the definition of a dataset: a data.frame, or an inherited class of it (data.table, [tibble](https://tibble.tidyverse.org/)); or a well-structured list (for example, a [json](https://arxiv.org/abs/1403.2805) object.)

## Managing citations and bibliographies {#bibliography}

```{r bibentry}
iris_bibentry <- bibentry_dataset(iris_dataset)
toBibtex(iris_bibentry)
```



```{r bibliography, results='asis'}
print(iris_bibentry, sytle="html")
```

## faIR: Interoperable & Reusable Datasets

The interoperability and reusability of datasets is greatly enhanced if they follow a standardized and practical format.  Our datasets follow the tidy data principles^[Wickham, H.(2014). Tidy Data. Journal of Statistical Software, 59(10), 1–23. <https://doi.org/10.18637/jss.v059.i10>] and are interoperable with the W3C^[DF Data Cube Vocabulary]^[RDF Data Cube Vocabulary, W3C Recommendation 16 January 2014  <https://www.w3.org/TR/vocab-data-cube/#metadata>] (semantic web) and SDMX (statistical) dataset definitions.

Both W3C and SDMX uses are more complex object, the Datacube in its description. The dataset is a reused datacube. To adhere to tidy data principles and easy use in reproducible research workflows, we further reduced our subjective definition of the dataset. 

- The `dataset` constructor first subsets the dataset for the `obs_id` observation identifier, and if it is missing, it creates one.
- Then it selects the `dimensions`, such as geographic concept or time concept. The iris dataset does not have these variables, so we do not select anything.
- Next we select the `measurements`.  In case only one `measurement` is present, we have a long-form dataset that can be easily serialized into an `RDF` object, for example. 
- Next we select any `attributes` that are unlikely to be used for statistical aggregation (unlike the dimensions) and which are not measured values. 
- We can pass on further optional dataset attributes. These attributes do not correspond with a single observation, rather the entire dataset.

```{r petalexample}
petal_length <- dataset(subset(iris, select = c("Petal.Length", "Species")), 
        dimensions = NULL, 
        measures   = "Petal.Length", 
        attributes = "Species")

petal_width <- dataset(subset(iris, select = c("Petal.Width", "Species")), 
        dimensions = NULL, 
        measures   = "Petal.Width", 
        attribute  = "Species")

require(dplyr)
petal_length %>%
  left_join (petal_width, by = c("Species")) %>%
  sample_n(10)
```

The obvious motivation of this format is that the datasets can be easily integrated, joined, combined, because they are tidy.
