---
title: "Datasets With FAIR Metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Datasets With FAIR Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupknitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dataset)
```

According to the [RDF Data Cube Vocabulary](https://www.w3.org/TR/vocab-data-cube/) DataSet is a collection of statistical data that corresponds to a defined structure. The data in a data set can be roughly described as belonging to one of the following kinds:

- `Observations`: these are the measured values, and the cells of a data frame object in R.
- `Organizational structure`: To locate an observation within the hypercube, one has at least to know the value of each dimension at which the observation is located, so these values must be specified for each observation. Datasets can have additional organizational structure in the form of slices as described in section 7.2.
- `Structural metadata`: Metadata to interpret the data. What is the unit of measurement? Is it a normal value or a series break? Is the value measured or estimated? These metadata are provided as attributes and can be attached to individual observations, or to higher levels.
- `Reference metadata`: Metadata that describes the dataset as a whole, such as categorization of the dataset, its publisher, or an endpoint where it can be accessed.

| Information | dataset |
|:--:|----|
| dimensions   |  first column section of the dataset   |
| measurements |  second column section of the dataset  |
| attributes   |  third column section of the dataset   |
| reference    |  attributes of the R object            |

## FAIR metadata

Our `dataset` R package aims to increase the *Findability, Accessibility, Interoperability,* and *Reuse* of digital assets, particularly datacubes and datasets used in statistics and data analysis. The [FAIR](https://www.go-fair.org/fair-principles/) principles  "…emphasise machine-actionability (i.e., the capacity of computational systems to find, access, interoperate, and reuse data with none or minimal human intervention) because humans increasingly rely on computational support to deal with data as a result of the increase in volume, complexity, and creation speed of data."

The `dataset` package adds metadata to R objects for full compatibility [The RDF Data Cube Vocabulary](https://www.w3.org/TR/vocab-data-cube/) [which includes the DataSet definition], which is harmonized with the [Statistical Data and Metadata eXchange](https://sdmx.org/).  This is necessary to correctly import data from statistical sources or from the semantic web, and harmonize the processed results with such services.

The [The RDF Data Cube Vocabulary](https://www.w3.org/TR/vocab-data-cube/) in itself uses some core elements fo the [Dublin Core Metadata Elements 1.1](https://www.dublincore.org/specifications/dublin-core/dces/), which is a library metadata standard to archive, store, and find information. The `dataset` R package goes beyond the DataSet definition and allows to add all the *Dublin Core Metadata Elements 1.1*, and to add the [DataCite Mandatory Properties 4.4](https://support.datacite.org/docs/datacite-metadata-schema-44) and the [DataCite Recommended and Optional Properties 4.3](https://support.datacite.org/docs/schema-optional-properties-v43).

The `Dublin Core` is an ISO Standard 15836:2009 since February 2009 [ISO15836]. Currently the 15 "core" elements are part of a bigger [DCMI Metadata Terms](https://www.dublincore.org/specifications/dublin-core/dcmi-terms/). The *DataCite* mandatory elements are narrower than the DublinCore elements, and the full DataCite is broader. The two standards are very-well harmonized with little difference.

We give a preference for *DataCite*, because it is a more modern standard that is better sutited to document software and data products. It is also the preferred choice of the EU open repositories that are used to deposit publicly funded research results, including datasets.


```{r}
iris_ds <- dataset(x=iris, 
                   Dimensions=NULL, 
                   Measures=c("Sepal.Length", "Sepal.Width", 
                              "Petal.Length", "Petal.Width"),
                   Attributes = "Species", 
                   Title = "Iris Dataset", 
                   Label = "The famous iris dataset used in R examples")

dublincore(iris_ds)
```
The `dataset()` constructor adds W3C/SDMX compatible structural metadata (declares the measured values and the attributes of the observations) and some Dublin Core data for findability (these are properties that are the same in DataCite.)

You can add more descriptive metadata to further support discovery, interoperability.  You do not need the `dataset()` class, any data.frame, i.e. `data.frame()`, `tibble::tibble()` or `data.table::data.table()` will do. The dataset constructor does not alter the data or the data structure of a data frame or an object inherited from data frame, only adds standardized metadata to it. 

```{r createdataset}
iris_ds <- datacite_add(iris,
                        Title = "Iris Dataset",
                        Creator = person(family ="Anderson", given ="Edgar", role = "aut"),
                        Identifier = "https://doi.org/10.1111/j.1469-1809.1936.tb02137.x",
                        Publisher= "American Iris Society",
                        PublicationYear = 1935,
                        Geolocation = "US",
                        Language = "en", 
                        Version = "1.0")

datacite(iris_ds)
```
```{r, message=FALSE, eval=FALSE}
library(data.table)
```
```{r, eval=FALSE}
datacite_add(x = data.table::data.table(iris), 
             Title = "Iris Dataset", 
             Creator =  person(family ="Anderson", given ="Edgar", role = "aut"))
```

Currently, the DataCite properties (all mandatory, and what was filled up from the optional ones) can be seen along with standard R metadata. As a property of the `dataset` class, which follows W3C and SDMX standards, and `survey`, which will follow DDI standards, the history of the `dataset` from creation (import) will be recorded as attribute metadata.

```{r showattributes}
dct_iris <- dublincore_add(
  x = iris,
  Title = "Iris Dataset",
  Creator = person("Edgar", "Anderson", role = "aut"),
  Publisher = "American Iris Society",
  Identifier = "https://doi.org/10.1111/j.1469-1809.1936.tb02137.x",
  Date = 1935,
  Language = "en"
)

dublincore(dct_iris)
```

## FAir: Findable & Accessible Datasets

If you work with R, you are almost certainly familiar with the iris dataset.  The *?iris* will provide you with some information about this often used dataset in tutorials. But how you make sure that you do not forget its important properties?

The function datacite [DataCite](https://datacite.org/) add at least the mandatory properties of the [DataCite Metadata Schema 4.3](https://schema.datacite.org/), a list of core metadata properties chosen for an accurate and consistent identification of a resource for citation and retrieval purposes. DataCite is largely interoperable to the other similar international standard, the [Dublin Core](https://www.dublincore.org/).  We will later add similar `dublincore` function, however, the practical differences are so small that adjustments, if needed, can be easily made by hand.

```{r datacite}
iris_dataset <- datacite_add(
  x = iris,
  Title = "Iris Dataset",
  Creator = person("Anderson", "Edgar", role = "aut"),
  Publisher= "American Iris Society",
  Identifier = "https://doi.org/10.1111/j.1469-1809.1936.tb02137.x",
  PublicationYear = 1935,
  Description = "This famous (Fisher's or Anderson's) iris data set gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica.",
  Language = "en")
```
```{r}
datacite(iris_dataset)
```
The `x` parameter can be any well-structured R object that meets the definition of a dataset: a data.frame, or an inherited class of it (data.table, [tibble](https://tibble.tidyverse.org/)); or a well-structured list (for example, a [json](https://arxiv.org/abs/1403.2805) object.)

## Managing citations and bibliographies {#bibliography}

```{r bibentry}
iris_bibentry <- bibentry_dataset(iris_dataset)
toBibtex(iris_bibentry)
```



```{r bibliography, results='asis'}
print(iris_bibentry, sytle="html")
```

## faIR: Interoperable & Reusable Datasets

The interoperability and reusability of datasets is greatly enhanced if they follow a standardized and practical format.  Our datasets follow the tidy data principles^[Wickham, H.(2014). Tidy Data. Journal of Statistical Software, 59(10), 1–23. <https://doi.org/10.18637/jss.v059.i10>] and are interoperable with the W3C^[DF Data Cube Vocabulary]^[RDF Data Cube Vocabulary, W3C Recommendation 16 January 2014  <https://www.w3.org/TR/vocab-data-cube/#metadata>] (semantic web) and SDMX (statistical) dataset definitions.

Both W3C and SDMX uses are more complex object, the Datacube in its description. The dataset is a reused datacube. To adhere to tidy data principles and easy use in reproducible research workflows, we further reduced our subjective definition of the dataset. 

- The `dataset` constructor first subsets the dataset for the `obs_id` observation identifier, and if it is missing, it creates one.
- Then it selects the `dimensions`, such as geographic concept or time concept. The iris dataset does not have these variables, so we do not select anything.
- Next we select the `measurements`.  In case only one `measurement` is present, we have a long-form dataset that can be easily serialized into an `RDF` object, for example. 
- Next we select any `attributes` that are unlikely to be used for statistical aggregation (unlike the dimensions) and which are not measured values. 
- We can pass on further optional dataset attributes. These attributes do not correspond with a single observation, rather the entire dataset.

```{r petalexample}
petal_length <- dataset(subset(iris, 
                               select = c("Petal.Length", "Species")), 
        Dimensions = NULL, 
        Measures   = "Petal.Length", 
        Attributes = "Species")

petal_width <- dataset(subset(iris, 
                              select = c("Petal.Width", "Species")), 
        Dimensions = NULL, 
        Measures   = "Petal.Width", 
        Attribute  = "Species")
```
```{r loaddplyr, message=FALSE}
require(dplyr)
```
```{r petalexampleprint}
petal_length %>%
  left_join (petal_width, by = c("Species")) %>%
  sample_n(10)
```

The obvious motivation of this format is that the datasets can be easily integrated, joined, combined, because they are tidy.
