---
title: "Data Provenance: Who Did What With the Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Provenance: Who Did What With the Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dataset)
```


The `dataset()` sub-class extends the data frames with various metadata, including provenance metadata.

```{r provenance}
provenance(iris_dataset)
```

Let's add the R programming language to the wasInformedBy field with the subject heading of the Library of Congress, which is easy to understand for humans and machines alike: [R (Computer program language)](https://id.loc.gov/authorities/subjects/sh2002004407.html)

```{r update-provenance}
provenance(iris_dataset) <- list( 
  wasInformedBy = "http://id.loc.gov/authorities/subjects/sh2002004407"
  )
```

Let us review the new provenance metadata:

```{r review-provenance, warning=FALSE}
provenance(iris_dataset)$wasInformedBy
```

Write it to triples:

```{r provenance-triples}
provenance_df <- as.data.frame(
  lapply(provenance(iris_dataset), function(x) x[[1]])
  )
provenance_df            <- id_to_column(provenance_df, "eg:dataset-1")
provenance_df$started_at <- xsd_convert(provenance_df$started_at)
provenance_df$ended_at   <- xsd_convert(provenance_df$ended_at)

dataset_to_triples(provenance_df, idcol="rowid")
```



## PROV

The _dataset_ package follows the PROV Data Model [[PROV-DM](https://www.w3.org/TR/prov-o/#bib-PROV-DM)] and expresses the provenance metadata using the PROV Ontology ([PROV-O](https://www.w3.org/TR/prov-o/)) expresses using the OWL2 Web Ontology Language (OWL2) [[OWL2-OVERVIEW](https://www.w3.org/TR/prov-o/#bib-OWL2-OVERVIEW)]. It provides a set of classes, properties, and restrictions that can be used to represent and interchange provenance information generated in different systems. The [PROV Document Overview](http://www.w3.org/TR/2012/WD-prov-overview-20121211/) describes the overall state of PROV, and should be read before other PROV documents.

