---
title: "Export And Publish a dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Export And Publish a dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dataset)
```

One of the primary motivations of the `dataset` and the `dublincore()` and `datacite()` functions is the creation of publication-ready data assets. 

## Exporting functions

The planned export functions will create `csv`, `xlsx`, SPSS `sav` and STATA `dta` datasets that contain all [FAIR](https://www.go-fair.org/fair-principles/) metadata.

## Publishing functions

We are planning an automatic publishing connection to [zen4R](https://github.com/eblondel/zen4R), the R language interface to the Zenodo open science repository. The `zen4R` package uses R6 classes which are known only for advanced R users. Our planned connector is a simple function that connects our s3 [dataset](), which is inherited from a simple `data.frame()` with the [depositRecord](https://github.com/eblondel/zen4R/wiki#ZenodoDeposit-publish) method of `zen4R` for immediate publication on Zenodo. 
