---
title: "From R to RDF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From R to RDF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupvignette, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (!requireNamespace("rdflib", quietly = TRUE)) {
  stop("Please install 'rdflib' to run this vignette.")
}

if (!requireNamespace("jsonld", quietly = TRUE)) {
  stop("Please install 'jsonld' to run this vignette.")
}
```

## From tidy data to RDF triples

This vignette demonstrates how to convert tidy R datasets into semantically enriched RDF triple structures, using the `dataset` and `rdflib` packages. These packages help you annotate variables with machine-readable concept definitions, units, and links to controlled vocabularies.

We’ll start with a small example of a tidy dataset representing countries (`geo`) with unique identifiers (`rowid`) and then show how to transform the dataset into RDF triples using standard vocabularies.

```{r setup}
library(dataset)
library(rdflib)
data("gdp")
```

## Creating a minimal semantically defined dataset

```{r smallgeo}
small_geo <- dataset_df(
  rowid = defined(
    paste0("eg:", seq_along(gdp$geo[1:3])),
    namespace = "http://example.com/dataset#"
  ),
  geo = defined(
    gdp$geo[1:3],
    label = "Geopolitical entity",
    concept = "http://purl.org/linked-data/sdmx/2009/dimension#refArea",
    namespace = "https://dd.eionet.europa.eu/vocabulary/eurostat/geo/$1"
  )
)

```

```{r triplesdf}
triples_df <- dataset_to_triples(small_geo)
knitr::kable(triples_df)
```

This produces triples like:

```{r createntriples}
ntriples <- dataset_to_triples(small_geo, format = "nt")
cat(ntriples, sep = "\n")
```

Each row of your dataset becomes a **subject**, each variable a **predicate**, and each value either a **URI** or a typed literal (like a date or number) — depending on how it's defined.

### RDF triples enable interoperability

The Resource Description Framework (RDF) represents data as subject–predicate–object triples. This allows your dataset to be machine-readable, linkable to external vocabularies, and queryable via SPARQL.

### RDF triples enable interoperability

The Resource Description Framework (RDF) represents data as subject–predicate–object triples. This allows your dataset to be machine-readable, linkable to external vocabularies, and queryable via SPARQL.

```{r ntripleexample}
n_triple(
  s = "http://example.com/dataset#eg:1",
  p = "http://purl.org/dc/terms/title",
  o = "Small Country Dataset"
)
```



```{r readrdf}
# We write to a temporary file our Ntriples created earlier
temp_file <- tempfile(fileext = ".nt")
writeLines(ntriples, con = temp_file)

rdf_graph <- rdf()
rdf_parse(rdf_graph, doc = temp_file, format = "ntriples")
rdf_graph
```
## Clean up

```{r clenup}
# Clean up: delete file and clear RDF graph
unlink(temp_file)
rm(rdf_graph)
gc()
```


## Scale up

```{r scaleup}
small_country_dataset <- dataset_df(
    geo = defined(
      gdp$geo,
      label = "Country name",
      concept = "http://dd.eionet.europa.eu/vocabulary/eurostat/geo/",
      namespace = "https://www.geonames.org/countries/$1/"
    ),
    year = defined(
      gdp$year,
      label = "Reference Period (Year)",
      concept = "http://purl.org/linked-data/sdmx/2009/dimension#refPeriod"
    ),
    gdp = defined(
      gdp$gdp,
      label = "Gross Domestic Product",
      unit = "CP_MEUR",
      concept = "http://data.europa.eu/83i/aa/GDP"
    ),
    unit = gdp$unit,
    freq = defined(
      gdp$freq,
      label = "Frequency",
      concept = "http://purl.org/linked-data/sdmx/2009/code"
    ),
    dataset_bibentry = dublincore(
      title = "Small Country Dataset",
      creator = person("Jane", "Doe"),
      publisher = "Example Inc.",
      datasource = "https://doi.org/10.2908/NAIDA_10_GDP",
      rights = "CC-BY",
      coverage = "Andorra, Lichtenstein and San Marino"
    )
)
```


```{r readrdf2}
# We write to a temporary file our Ntriples created earlier
temp_file <- tempfile(fileext = ".nt")
writeLines(dataset_to_triples(small_country_dataset, format = "nt"),
           con = temp_file)

rdf_graph <- rdf()
rdf_parse(rdf_graph, doc = temp_file, format = "ntriples")
rdf_graph
```
```{r readjsonld}
# Create temporary JSON-LD output file
jsonld_file <- tempfile(fileext = ".jsonld")

# Serialize to JSON-LD
rdf_serialize(rdf_graph, doc = jsonld_file, format = "jsonld")

# Read it back in R for display
# Only first 30 lines are shown...
cat(readLines(jsonld_file)[1:30], sep = "\n")
```

```{r clenup2}
unlink(temp_file)
rm(rdf_graph)
gc()
```
