---
title: "From dataset To RDF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From dataset To RDF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupRDFvignette, include = FALSE}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  message  = FALSE,
  comment = "#>",
  out.width = '90%'
)
if (file.exists(file.path(tempdir(), "example_ttl.ttl"))) {
  file.remove(file.path(tempdir(), "example_ttl.ttl"))
}

library(here)
```

Long-form triples are tidy datasets with an explicit row (observation) identifier among the columns.



&nbsp;          | &nbsp;  | &nbsp;      | &nbsp;
----------------|---------|-------------|-------------
**JSON**        | object  | property    | value
**spreadsheet** | row id  | column name | cell
**data.frame**  | key     | variable    | measurement
**data.frame**  | key     | attribute   | value
**RDF**         | subject | predicate   | object

Table source: [rdflib](https://cran.r-project.org/package=rdflib/vignettes/rdf_intro.html)

```{r setup}
library(dataset)
```

Let's take a small subset of the `iris_dataset()`, which is the semantically enriched version of the base R `iris` dataset. Limiting the the dataset to the top 3 rows, we have exactly 2 x 5 = 10 data cells.

```{r head-iris}
head(iris_dataset, 2)
```

```{r head-iris-xsd}
xsd_convert(head(iris_dataset, 2))
```

Let us arrange this to subject-predicate-object triples.

```{r dataset-to-triples}
iris_triples <- dataset_to_triples(xsd_convert(head(iris_dataset,2)))
iris_triples
```

We receive 2x5 = 10 rows; each with an identifier. The identifiers are made from `row.names()`, and we have exactly 5 statements about the first observation (`iris:o1`), and 5 statements about the second (`iris:o2`). Each statement simply states the observed value.


```{r define-vars}
iris_triples$p <- paste0("iris:", iris_triples$p)
iris_triples
```



```{r row-names}
row.names(head(iris_dataset,2))
```
```{r write-to-ttle}
vignette_temp_file <- file.path(tempdir(), "example_ttl.ttl")
dataset_ttl_write(dataset_to_triples(iris_triples), 
                  file_path = vignette_temp_file)
```


We see a standard metadata file expressed in the [Turtle](https://www.w3.org/TR/rdf12-turtle/) language. The definitions are
separated with a `# -- Observations ------` comment from the actual statements
about the dataset.

```{r read-ttl}
# Only first 23 lines are read and printed:
readLines(vignette_temp_file, n = 23)
```
If we would try to parse this file with a ttl-reader, we would get an error message, because not all statements are well-defined. 

## The prefix

The Turtle prefix statements define the abbreviations of the following namespaces:

```{r prefix-5}
readLines(vignette_temp_file, n = 5)
```
- [owl](https://www.w3.org/OWL/): the Web Ontology Language (OWL); which is the metadata language definition in which the rest of the definitions (vocabularies) are described.
- [rdf](http://www.w3.org/rdf/): Resource Description Framework (RDF). RDF provides the data model explaining how to build a graph. 
- [rdfs](http://www.w3.org/rdf/): RDFS is a vocabulary, in RDF, that explains how nodes of a graph relate. 
- [qb](https://www.w3.org/TR/vocab-data-cube/): The RDF Data Cube Vocabulary is the semantic definition of statistical datasets and their multi-dimensional versions, the datacubes, using `owl`, `rdf` and `rdfs`.
- [xsd](https://www.w3.org/TR/xmlschema11-1/): The W3C XML Schema Definition Language (XSD), which offers facilities for describing the structure and constraining the contents of XML documents. It ensures that R types, such as `character`, `integer`,  or `Date` are correctly represented in web documents.

The prefix makes the `ttl` Turle-file future-proof: before explaining the semantics of the data, it contains all the definitions that are needed to understand the explanation. It is a dictionary; every elements of the vocabulary that are needed to explain the iris dataset should be here.  This means that we must define the `iris` prefix, too.

These definitions can be found in the `data("dataset_namespace")` dataset.  we only need to add the definitions ourselves that is unique about our own dataset, in this case, the definitions of the variables of the iris dataset, i.e., the `iris` namespace:

```{r further-prefix}
data("dataset_namespace")
unique(get_prefix(row.names(head(iris_dataset,2))))
```

The `dataset_namespace` data file contains some often used vocabularies and their prefixes. Let us select  `owl:`, `rdf:`, `rdfs:`, `qb:` and add `iris:` as `<<www.example.com/iris#>>` (the _example.com_ domain is reserved by the World Wide Web consortium for documentation and tutorial examples.)

```{r define-all-prefixes}
used_prefixes <- which(dataset_namespace$prefix %in% c(
  "owl:", "rdf:", "rdfs:", "qb:", "xsd:")
  )

vignette_namespace <- rbind(
  dataset_namespace[used_prefixes, ], 
       data.frame (prefix = "iris:", 
                   uri = '<www.example.com/iris#>')
) 

vignette_namespace
```

Let us overwrite the earlier ttl file, but this time defining the variables and observations with the `iris:` prefix:

```{r write-to-ttle2}
dataset_ttl_write(
  iris_triples, 
  ttl_namespace = vignette_namespace,
  file_path = vignette_temp_file, 
  overwrite = TRUE)
```

```{r prefix2}
readLines(vignette_temp_file, n = 23)
```
## Working with rdflib

[RDFLib](https://rdflib.readthedocs.io/en/stable/) is a pure Python package for working with RDF with RDF serialisation parsers, store implementations, graph interface and a SPARQL query and update implementation. It has an excellent R binding, the [rdflib](https://cran.r-project.org/package=rdflib/vignettes/rdf_intro.html) package^[Carl Boettiger: [A tidyverse lover’s intro to RDF](https://cran.r-project.org/package=rdflib/vignettes/rdf_intro.html)].

In this section we show how to work further with our future-proof datasets. We parse the `ttl` file created with the dataset package into a triplestore:

```{r parse-ttl}
require(rdflib)
example_rdf <- rdf_parse(vignette_temp_file, format = "turtle")
example_rdf
```
And define a simple SPARQL query on the data:


```{r sparql_example}
sparql <-
'PREFIX iris: <www.example.com/iris#> 
 SELECT ?observation ?value
 WHERE { ?observation iris:Sepal.Length ?value . }'

rdf_query(example_rdf, sparql)
```

Convert, for example, to `JSON-LD` format...:

```{r convert-to-jsonld}
temp_jsonld_file <- file.path(tempdir(), "example_jsonld.json")
rdf_serialize(rdf=example_rdf, doc = temp_jsonld_file, format = "jsonld")
```

... and read in the first 12 lines:

```{r read-first-12-lines}
readLines(temp_jsonld_file, 12)
```

