---
title: "From dataset To RDF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From dataset To RDF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupRDFvignette, include = FALSE}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  message  = FALSE,
  comment = "#>",
  out.width = '90%'
)
library(here)
library(readxl)
library(kableExtra)
```

```{r setup}
library(dataset)
```

> This documentation is not updated yet to the development version of the [dataset] package.

Our datasets are defined in a way that their dimensions can be easily and unambiguously reduced to triples for RDF applications; they can be easily serialized to, or synchronized with semantic web applications with the [rdflib](https://cran.r-project.org/package=rdflib/vignettes/rdf_intro.html) package^[Carl Boettiger: [A tidyverse lover’s intro to RDF](https://cran.r-project.org/package=rdflib/vignettes/rdf_intro.html)].

Read more about how we adopted to datacube model of SDMX and the [RDF Data Cube Vocabulary](https://www.w3.org/TR/vocab-data-cube/) for datasets as R objects in the [The dataset S3 Class](https://dataset.dataobservatory.eu/articles/RDF.html) vignette article.

Because our datasets conform the tidy data concept, they can be reduced into long-form triples.



&nbsp;          | &nbsp;  | &nbsp;      | &nbsp;
----------------|---------|-------------|-------------
**RDF**         | subject | predicate   | object
**JSON**        | object  | property    | value
**spreadsheet** | row id  | column name | cell
**data.frame**  | key     | variable    | measurement
**data.frame**  | key     | attribute   | value


Table source: [rdflib](https://cran.r-project.org/package=rdflib/vignettes/rdf_intro.html)

## Dimension reductions of the dataset

Our datasets are tidy.

```{r importexamplexlsx}
example_df <- readxl::read_excel(
  system.file("extdata", "rdf_example.xlsx", package = "dataset"),
  sheet = "dataset-wide")
```

