---
title: "From R to RDF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From R to RDF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setupvignette, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

if (!requireNamespace("rdflib", quietly = TRUE)) {
  stop("Please install 'rdflib' to run this vignette.")
}

if (!requireNamespace("jsonld", quietly = TRUE)) {
  stop("Please install 'jsonld' to run this vignette.")
}
```

```{css, echo = FALSE}
    .small .table{
      font-size: 10px;
    }
```

## From tidy data to RDF triples

This vignette demonstrates how to convert tidy R datasets into semantically enriched RDF triple structures, using the `dataset` and `rdflib` packages. These packages help you annotate variables with machine-readable concept definitions, units, and links to controlled vocabularies.

We’ll start with a small example of a tidy dataset representing countries (`geo`) with unique identifiers (`rowid`) and then show how to transform the dataset into RDF triples using standard vocabularies.

```{r setup}
library(dataset)
library(rdflib)
data("gdp")
```

## Creating a minimal semantically defined dataset

```{r smallgeo}
small_geo <- dataset_df(
  rowid = defined(
    paste0("eg:", seq_along(gdp$geo[1:3])),
    namespace = "http://example.com/dataset#"
  ),
  geo = defined(
    gdp$geo[1:3],
    label = "Geopolitical entity",
    concept = "http://purl.org/linked-data/sdmx/2009/dimension#refArea",
    namespace = "https://dd.eionet.europa.eu/vocabulary/eurostat/geo/$1"
  )
)
```

::: small
```{r triplesdf}
triples_df <- dataset_to_triples(small_geo)
knitr::kable(triples_df)
```
:::

This produces triples like:

::: small
```{r createntriples}
ntriples <- dataset_to_triples(small_geo, format = "nt")
cat(ntriples, sep = "\n")
```
:::

Each row of your dataset becomes a **subject**, each variable a **predicate**, and each value either a **URI** or a typed literal (like a date or number) — depending on how it's defined.

### RDF triples enable interoperability

The *Resource Description Framework* (RDF) represents data as subject–predicate–object triples. This allows your dataset to be machine-readable, linkable to external vocabularies, and queryable via SPARQL. The dataset is presented cell by cell as an elementary semantic (knowledge) statement, carrying the definitions of the observed unit, the measurement or attribute, and the relationship between the observed unit and this variable. Instead of carrying a label like "GDP", or "Gross Domestic Product", it has a precise definition for this statistical variable that anybody can look up.

```{r ntripleexample}
one_triple <- n_triple(
  s = "http://example.com/dataset#eg:1",
  p = "http://purl.org/dc/terms/title",
  o = "Small Country Dataset"
)
```

::: small
```{r}
cat(one_triple)
```
:::

```{r readrdf}
# We write to a temporary file our N-triples created earlier
# Initialise a connection, this time in tempfile().
temp_file <- tempfile(fileext = ".nt")

# Write the text into the connection:
writeLines(ntriples, con = temp_file)

# Initialise an RDF graph in memory:
rdf_graph <- rdf()
```

Read the contents of the temporary text file into the graph:

::: small
```{r showgraph}
rdf_parse(rdf_graph, doc = temp_file, format = "ntriples")
```
:::

## Clean up

It is a good practice to close connections, or clean up larger objects living in the memory:

```{r clenup}
# Clean up: delete file and clear RDF graph
unlink(temp_file)
rm(rdf_graph)
gc()
```

## Scale up

```{r scaleup}
small_country_dataset <- dataset_df(
  geo = defined(
    gdp$geo,
    label = "Country name",
    concept = "http://dd.eionet.europa.eu/vocabulary/eurostat/geo/",
    namespace = "https://www.geonames.org/countries/$1/"
  ),
  year = defined(
    gdp$year,
    label = "Reference Period (Year)",
    concept = "http://purl.org/linked-data/sdmx/2009/dimension#refPeriod"
  ),
  gdp = defined(
    gdp$gdp,
    label = "Gross Domestic Product",
    unit = "CP_MEUR",
    concept = "http://data.europa.eu/83i/aa/GDP"
  ),
  unit = gdp$unit,
  freq = defined(
    gdp$freq,
    label = "Frequency",
    concept = "http://purl.org/linked-data/sdmx/2009/code",
    namespace = "http://purl.org/linked-data/sdmx/2009/code#freq-"
  ),
  dataset_bibentry = dublincore(
    title = "Small Country Dataset",
    creator = person("Jane", "Doe"),
    publisher = "Example Inc.",
    datasource = "https://doi.org/10.2908/NAIDA_10_GDP",
    rights = "CC-BY",
    coverage = "Andorra, Lichtenstein and San Marino"
  )
)
```

```{r smallcountrydfnt}
small_country_df_nt <- dataset_to_triples(small_country_dataset,
  format = "nt"
)
```

The following lines read as:

-   [1] `Observation #1` is a geopolitical entity, `Andorra`.
-   [11] `Observation #1` has a reference time period of `2020`.
-   [21] `Observation #1` has a decimal GDP value of `2354.8`
-   [31] `Observation #1` has a unit of `million euros, current prices`.
-   [41] `Observation #1` has a measurement frequency that is `annual`.

```{r smallcountrydfntsample}
## See rows 1,11,21
small_country_df_nt[c(1, 11, 21, 31, 41)]
```

The statements about `Observation 1`, which is Andorra's national economy in 2020, is not serialised consecutively in the text file. This is not necessary, because each cell is precisely connected to the *row* (first part of the triple) and *column* (second part of the triple). We could say that the entire map to the original dataset is embedded into the flat text file, therefore it can be easily imported into a database.

```{r readrdf2}
# We write to a temporary file our Ntriples created earlier
temp_file <- tempfile(fileext = ".nt")
writeLines(small_country_df_nt,
  con = temp_file
)

rdf_graph <- rdf()
rdf_parse(rdf_graph, doc = temp_file, format = "ntriples")
rdf_graph
```

Your dataset is now ready to be exported to meet the true FAIR standards, because they are:

- **self-descriptive**: variables carry labels, units, and definitions.
- **machine-readable**: linked vocabularies and standard identifiers.
- **ready to publish and share**: they carry the metadata of each variable, potentially each observation unit, and through metadata standards like Dublin Core and DataCite the information about the whole dataset, too.

```{r readjsonld}
# Create temporary JSON-LD output file
jsonld_file <- tempfile(fileext = ".jsonld")

# Serialize (export) the entire graph to JSON-LD format
rdf_serialize(rdf_graph, doc = jsonld_file, format = "jsonld")
```

Read it back to R for display (only first 30 lines are shown):

:::: small
```{r readjsonldprint}
cat(readLines(jsonld_file)[1:30], sep = "\n")
```
::::

```{r clenup2, echo=FALSE, message=FALSE}
unlink(temp_file)
rm(rdf_graph)
gc()
```


