---
title: "The survey Class"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The survey Class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dataset)
library(declared)
```

You need the latest development version of [declared](https://github.com/dusadrian/declared).

```{r, eval=FALSE}
remotes::install_github('dusadrian/declared')
```


The survey class will be derived from the `dataset` class. 

```{r}
obs_id <- c("Saschia Iemand", "Jane Doe", "Jack Doe", "Pim Iemand", "Matti Virtanen" )
sex <- declared ( c(1,1,0,-1,1), 
                  labels = c(Male = 0, Female = 1, DK = -1), 
                  na_values = -1)
geo <- c("NL-ZH", "IE-05", "GB-NIR", "NL-ZH", "FI1C")
```

```{r}
difficulty_bills <- declared (
  c(0,1,2,-1,0), 
  labels = c(Never = 0,  Time_to_time = 1, Always = 2, DK = -1)
)
age_exact <- declared (
  c( 34,45,21,55,-1), 
  labels = c( A = 34,A = 45,A  = 21, A= 55, DK = -1)
)
listen_spotify <- declared (
  c(0,1,9,0,1),
  labels = c( No = 0, Yes = 1,Inap = 9), 
  na_values = 9
)
```

```{r}
raw_survey <- data.frame ( 
  obs_id = obs_id, 
  geo = geo, 
  listen_spotify = listen_spotify,
  sex = sex,
  age_exact = age_exact, 
  difficulty_bills = difficulty_bills
)

survey_dataset  <- dataset( x= raw_survey,
                            dimensions = "geo", 
                            measures = c("listen_spotify", "sex",
                                         "age_exact", "difficulty_bills"), 
                            attributes = NULL, 
                            sdmx_attributes = "geo", 
                            Title = "Tiny Survey", 
                            Creator = person("Jane", "Doe"))
```

```{r}
datacite(survey_dataset)
```

It is a good practice to define valid, but not present labels in `declared`, because in the retrospective harmonization workflow they may be concatenaded (binded) together with further observations that do have the currently not used label. 

In this example, the `DK` or declined label is not in use.

```{r, eval=FALSE}
# This is not valied in declared
listen_spotify <- declared(
  c(0,1,9,0,1),
  labels = c( No = 0, Yes = 1,Inap = 9, DK =-1), 
  na_values = c(9, -1)
  )
```



```{r}
print(listen_spotify)
```


```{r}
c(listen_spotify, declared(
  c(-1,-1,-1),
  labels = c( No = 0, Yes = 1,Inap = 9, DK =-1)
  ))
```

```{r}
summary(listen_spotify)
```





```{r}
survey_dataset <- datacite_add(survey_dataset, 
                               Title = "Tiny Survey", 
                               Creator = person("Daniel", "Antal"), 
                               Identifier = "https://doi.org/xxxx.yyyyy",
                               Publisher = "Reprex", 
                               PublicationYear = 2022, 
                               Subject = "Surveys", 
                               Language = "en")
```

The `survey` class inherits elements of the `dataset` class, but it will be more strictly defined.  I am considering to make  `declared` every single column except for the `obs_id`. Even `numeric` types with `Inap` and `DK` would map nicely to `CL_OBS_STATUS` SDMX codes that make missing observation explicit, and try to categorize them.

```{r}
datacite(survey_dataset)
```

Is the `summary` method implemented for `declared`? Both `dataset` and `survey` will need new `print` and `summary` methods.

```{r}
summary(survey_dataset)
```
The `survey` (should) contain the entire processing history from creation, and optionally the `DataCite` schema for publication created with `datacite_add()`.  A similar `dublincore_add` function uses the Dublin Core metadata definitions.

Eventually, a connection to the packages [zen4R](https://github.com/eblondel/zen4R) will make sure that the correctly described dataset can get a Zenodo record, receive a DOI, the DOI recorded in the object, and upload to Zenodo.
